---
title: "`canaper`: Categorical analysis of neo- and paleo-endemism in R"
author:
  - name: Joel H. Nitta
    affil: 1
    orcid: '0000-0003-4719-7472'
affiliation:
  - num: 1
    address: "University of Tokyo, Department of Biological Sciences<br>✉️ joelnitta@gmail.com<br>🌍 https://joelnitta.com<p style='font-size:60px'>Check out the repo: https://github.com/joelnitta/canaper</p>"
column_numbers: 3
logoright_name: "images/logo_ut.png"
output: 
  posterdown::posterdown_html:
    self_contained: false
    template: template.html
csl: elsevier-with-titles.csl
bibliography: references.yaml
---

```{css, style, echo = FALSE}
.hanging div.csl-entry {
  margin-left:2em;
  text-indent:-2em;
}
div.csl-left-margin {
  min-width:2em;
  float:left;
}
div.csl-right-inline {
  margin-left:2em;
  padding-left:1em;
}
div.csl-indent {
  margin-left: 2em;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
#title code {
  color: inherit;
  font-size: inherit;
  background-color: transparent;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
options(digits = 2)
library(canaper)

set.seed(071221)
```

# 1. Background

* CANAPE (**C**ategorical **A**nalysis of **N**eo- and **P**aleo-endemism) is a recently developed method that provides insight into the evolutionary processes underlying endemism [@Mishler2014]
* CANAPE integrates a phylogenetic tree with a community (species × sites) matrix to infer if endemic areas are so because of recent speciation (neoendemism), or because they harbor old lineages that have mostly gone extinct in other areas (paleoendemism), or if they are a mixture of both
* CANAPE is currently only implemented in Biodiverse [@Laffan2010], a program written in perl that is used via a GUI or custom scripts.
* <span style='font-size:120%'>**`canaper` is a new R package to conduct CANAPE entirely in R**</span>

# 2. Features

* Sparse matrix encoding of community matrices to increase computing efficiency via `phyloregion` [@Daru2020a]
* Simple implementation of parallel computing to increase speed via `future`

# 3. Installation and loading

`canaper` is currently available on GitHub.

```{r install, eval = FALSE}
devtools::install_github("joelnitta/canaper")
```

```{r load-pkg}
# Load packages
library(canaper)
library(ape) # For handling phylogenies
library(future) # For parallel computing
library(tidyverse) # For data wrangling and visualization
```

# 4. Example: Australian *Acacia*

This demonstrates the package using the dataset of the paper where CANAPE was first published [@Mishler2014]: `r ncol(acacia$comm)` species of *Acacia* in Australia distributed over `r nrow(acacia$comm)` sites:

```{r load}
# Phylogenetic tree
acacia$phy

# Community data matrix (in part). 
# Rownames correspond to geographical coordinates
acacia$comm[1:4, 1:4]
```
<br>

## 4.1. Run randomization test

The first step of CANAPE is to compare the observed values of phylogenetic endemism (PE) and alternative PE (PE measured on a modified tree where all branch lengths are set equal) with those from a set of random communities. The `cpr_rand_test()` conducts the randomization, using parallel computing to increase speed. The `picante` package [@Kembel2010] is used to generate the random communities.

```{r rand-test}
# Set a parallel back-end, with 4 CPUs running simultaneously
plan(multisession, workers = 4)

# Run randomization test
acacia_rand_res <- cpr_rand_test(
  acacia$comm, acacia$phy, 
  n_reps = 100, n_iterations = 100000)

# Check some of the results
acacia_rand_res %>%
  slice(1:3) %>%
  select(pe_obs, pe_rand_mean, pe_obs_p_upper, pe_obs_p_lower) %>%
  as_tibble()
```

Output summary (in part):

* `pe_obs` = observed PE
* `pe_rand_mean` = mean PE of the randomizations
* `pe_obs_p_upper` = percent of randomizations where observed PE was greater than random values
* `pe_obs_p_lower` = percent of randomizations where observed PE was lower than random values

## 4.2. Classify significance

The next step of CANAPE is to classify the endemism type of each site. The `cpr_classify_endem()` function does this automatically given output of `cpr_rand_test()`.

```{r classify}
# Classify endemism type
acacia_canape <- cpr_classify_endem(acacia_rand_res)
# Count the results
count(acacia_canape, endem_type)
```

Endemism codes:

* `neo` = Neo-endemic
* `paleo` = Paleo-endemic
* `mixed` = Mix of neo and paleo
* `super` = Mixed, and highly significant (*p* < 0.01)

<br>

## 4.3. Visualize results

We can visualize the results of CANAPE using `ggplot2`. The `cpr_endem_cols` palette that is accessible regardless of color vision deficiency is provided.

```{r vis, fig.width = 10, fig.height = 8, fig.align = "center"}
# Fist do some data wrangling to make 
# the results easier to plot (add lat/long columns)
acacia_canape <- as_tibble(acacia_canape, rownames = "site") %>% 
  separate(site, c("lon", "lat"), sep = ":") %>% 
  mutate(dplyr::across(c(lon, lat), parse_number))

# Plot the results
ggplot(acacia_canape, aes(x = lon, y = lat, fill = endem_type)) + 
  geom_tile() +
  scale_fill_manual(values = cpr_endem_cols, name = "Endemism type") +
  guides(fill = guide_legend(title.position = "top")) +
  theme_gray(base_size = 24) +
  theme(legend.position = "bottom")
```

# 5. Next steps

* Implement `rand_structured` randomization algorithm of Biodiverse (should speed up randomizations ~10×)
* Submit to CRAN

# Acknowledgements

Thanks to Shawn Laffan for providing help with the code.

# References

<span style='font-size:60%'><div id="refs"></div></span> 

<span style='font-size:100%'>Source code: https://github.com/joelnitta/botany_poster_2021</span> 
